Visualizing Contingency Tables

### 2007-10-02 trunk/vcd/paper/Handbook/greyscale/hsc.Rnw
grey palette version added
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@657 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

set.seed(myseed)
hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))
@

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001} or spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}.
%% or spinograms \cite{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

@@

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the frequency of visits (from relatives or friends). 
The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
%FIXME: lookup regular in Wing-paper.
patients that were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data 
that the longer the length of stay in hospital, the less frequent the
visits, which can be seen from the column-standardized table 
(see Table~\ref{tab:colhospital}).

@@

\index{mosaic plot} Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
Chapter III-12 in this book provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. \index{spine plot}
From this plot, we see that the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Figure~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
to be the explanatory variable.

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If one variable is
explanatory, it should be be used first for splitting: thus, the display shows
the conditional distribution of the dependent variable given the explanatory one.
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models are of primary interest.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<color,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = diverge_hcl(5, c = 100, l = c(50, 90)))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = diverge_hsv(5))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@ 
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for 
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.
  In a monochrome version of the paper, the right and left hand
  sides of the HCL colors palette appear identical. }
\label{fig:diverging}
\end{center}
\end{figure}

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
\index{color space!CIELAB} \index{color space!CIELUV}
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
\index{color space!HCL}
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminance better 
contrasts can be achieved.

@@

All plots introduced in Section~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with
highlighting \index{highlighting}
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
\index{residual-based shading} residual-based shading
of the tiles to additionally visualize the
residuals from a given independence model fitted to the table. 
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively. 
The heuristic for choosing the cut-offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
Friendly shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}. 
In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above, useful when no color can be used.

In Figure~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

@@

For the visualization of areas perceptually uniform colors and color palettes
should always be used and HCL color space provides a convenient way to do so.
Shadings can be used to add information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis of relationships with a 
single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
by using diverging color palettes to code the sign and size of a residual.
In addition, the significance of test statistics can be visualized by using 
overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
statistic(s) allows for data-driven cut-offs in the diverging palettes and
precise diagnostic identification of the cells in conflict with the
null hypothesis of independence.

@@

In Section~\ref{sec:twoway}, we have presented basic displays
for two-way tables, based on the visualization of information on 
the tables' cells, arranged in rectangular form. For multi-way tables,
mosaic plots can directly be used by simply adding further splits for each 
additional variable. For sieve and association plots,
we apply the basic idea of mosaic plots to the table itself,
i.e., simply nest the variables into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation of the multi-way table that can be visualized in %'
ways similar to a two-dimensional table.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that is to be ``explained'' by the others. 
A natural way of representing such a three-way table is to use a mosaic
display, 
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the 
\index{doubledecker plot} doubledecker plot 
in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F and in between for C to E). The rates 
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

One approach is employing pairs plots (such as that in Figure~\ref{fig:pairs}) for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual 
% independence in the upper triangle, 
% mosaics for \emph{joint} independence 
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately 
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color 
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level 
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

>>>>

set.seed(myseed)
hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))

## greyscale-Function
togrey <- function(x) {
  x <- as(hex2RGB(substr(x, 1, 7)), "polarLUV")
  x@coords[,2] <- 0
  hex(x)
}
@

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001} or spine
  plots \cite{vcd:riedwyl+schuepbach:1994}.
%% or spinograms \cite{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2006b}.
\end{itemize}

@@

\noindent The table relates the length of stay (in years) of 132
long-term schizophrenic patients in two London mental hospitals with
the frequency of visits (from relatives or friends).
The length of stay (LOS) has been categorized into
2--9 years, 10--19 years, and more than 19 years. There are also three
categories for the visit frequency: regular (including
patients who were allowed to go home), less than monthly, and never.
Wing \cite{vcd:wing:1962} concludes from this data
that the longer the patients stayed in hospital, the less frequent they
are visited, which can be seen from the column-standardized table
(see Table~\ref{tab:colhospital}).

@@

\index{mosaic plot} Mosaic displays have been introduced by
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}.
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots;
Chapter III-12 in this book provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}).
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also
called a \emph{spine plot} \cite{vcd:hummel:1996}. \index{spine plot}
From this plot, we see that the number of patients
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different
compared to the previous mosaic.
In this alternative display (see Figure~\ref{fig:mosaic3fig}),
we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
to be the explanatory variable.

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If one variable is
explanatory, it should be be used first for splitting;
the display then shows the conditional distribution of the
dependent variable given the explanatory one.
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models are of primary interest.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<color,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = togrey(rainbow(9)), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = togrey(rainbow_hcl(9)), labels = 360 * 0:8/9)
@
\caption{Qualitative color palette for the HSV (left)
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., they will
         appear to be virtually identical.}
\label{fig:colors}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<diverging,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
par(mfrow = c(1,1), mar = c(1,1,1,1), oma = c(0,0,0,0))
plot.new()
rect(0:4/5, 0.2, 1:5/5, 0.5, border = 0, col = togrey(diverge_hcl(5, c = 100, l = c(50, 90))))
rect(0, 0.2, 1, 0.5, border = 1, col = NULL)
rect(0:4/5, 0.55, 1:5/5, 0.85, border = 0, col = togrey(diverge_hsv(5)))
rect(0, 0.55, 1, 0.85, border = 1, col = NULL)
text(c(1:5/5 - 0.1), 0.11, c("(260, 100, 50)", "(260, 50, 70)", "(H, 0, 90)", "(0, 50, 70)", "(0, 100, 50)"))
text(c(1:5/5 - 0.1), 0.91, c("(240, 100, 100)", "(240, 50, 100)", "(H, 0, 100)", "(0, 50, 100)","(0, 100, 100)"))
@
\caption{Diverging color palettes for the HSV space (upper part) and
  the HCL space (lower part), ranging from blue over a neutral color to
  red. The triples indicate the settings for
  the three dimensions: (Hue, Saturation, Value)
  in the upper part, and (Hue, Chroma, Luminance) in the lower part.
  In a monochrome version of the paper, the right and left hand
  sides of the HCL colors palette appear to be identical, unlike the
  HSV color palette. }
\label{fig:diverging}
\end{center}
\end{figure}

The use of colors that are more `in harmony' goes back to Munsell
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
\index{color space!CIELAB} \index{color space!CIELUV}
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
\index{color space!HCL}
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance
are always balanced towards the same grey
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right).
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2007} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}.
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue)
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter
full chroma colors), however, by varying both chroma and luminance better
contrasts can be achieved.

@@

The mosaic plots introduced in Section~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with
highlighting \index{highlighting}
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
\index{residual-based shading} residual-based shading
of the tiles to additionally visualize the
residuals from a given independence model fitted to the table.
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively.
The heuristic for choosing the cut-offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of this
shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}.
In addition to the shading of the rectangles themselves, the Friendly
shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above.

In Figure~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE,
       mar = c(left = 3.5),
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5),
         offset_varnames = c(left = 1, top = 0.5), set_labels =
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited patients,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

@@

For the visualization of areas perceptually uniform colors and color palettes
should always be used and HCL color space provides a convenient way to do so.
Shadings can be used to add information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis of relationships with a
single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
by using diverging color palettes to code the sign and size of a residual.
In addition, the non-significance of test statistics can be visualized by using
overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
statistic(s) allows for data-driven cut-offs in the diverging palettes and
precise diagnostic identification of the cells in conflict with the
null hypothesis of independence.

@@

In Section~\ref{sec:twoway}, we have presented basic displays
for two-way tables, based on the visualization of information on
the tables' cells, arranged in rectangular form. For multi-way tables,
mosaic plots can directly be used by simply adding further splits for each
additional variable. For sieve and association plots,
we apply the basic idea of mosaic plots to the table itself,
i.e., simply nest the variables into rows and columns using recursive
conditional splits, given the margins. The result is a
`flat' representation of the multi-way table that can be visualized in %'
a way similar to a two-dimensional table.

@@

The next step is to investigate three-way interactions among the variables.
In this example, we have a binary variable of interest (admission)
that is to be ``explained'' by the others.
A natural way of representing such a three-way table is to use a mosaic
display,
splitting first in the explanatory variables department and gender and highlighting
the resulting mosaic with respect to the dependent variable admissions. Here, we
use the vertical splits for both explanatory variables, resulting in the
\index{doubledecker plot} doubledecker plot
in Figure~\ref{fig:doubleucb}. From the widths of the
tiles it can be seen that students apply to the six departments in
unequal numbers (with a particularly small
number of females in departments~A and B),
and from the highlighting we see again that the admission rates differ
among the departments (roughly speaking, the admission rate is high for A and B,
low for F, and in between for C to E). The rates
are equal for male and female students, except for department A where
\textit{more} female than male students are admitted.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbhigh,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCB admissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

One approach is employing pairs plots (such as in Figure~\ref{fig:pairs}) for model search:
we can use the position of the cells in the pairs matrix to select an
independence model and add a corresponding shading to the tiles to
visualize the residuals.
More precisely, each cell $a_{ij}$ in such a matrix defines two
variables $i$ and $j$ that can be used to specify the model visualized
in that cell. Typical hypotheses are: Variables $i$ and $j$ are
marginally independent; variables $i$ and $j$ are conditionally
independent, given all others; variables $i$ and $j$ are jointly
independent from all others.
% Figure~\ref{fig:pairs} shows a pairs display with mosaic plots visualizing mutual
% independence in the upper triangle,
% mosaics for \emph{joint} independence
% in the lower triangle, and bar charts on the diagonal.
% In the upper part of the matrix, we can immediately
% detect the independence of hair and gender, and eye and gender from the
% non-colored cells in the corresponding mosaic plots. On the other hand, hair and eye color
% clearly are associated. In the lower part of the matrix, the mosaic
% at the intersection of hair and eye is the less colored one. Its shading
% represents the model of hair and eye color being jointly independent from
% gender, i.e., their association being homogeneous over the gender
% strata. It is the model that fits best, although still rejected at the
% 5\% level
% ($p=\Sexpr{round(summary(loglm(~ Hair * Eye + Sex, data = HairEyeColor))$tests[2,3],4)}$): %$
% possibly, because too many brown-haired male
% students have blue eyes, and too few of them have brown eyes, than expected under
% the null model?

Another approach is to visualize (the deviations from) a particular fitted model
with ideas similar to those described in Section~\ref{sec:extensions}.
Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression.
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A.
Therefore, a sensible way of representing the table is to
nest admission into department instead of gender to obtain a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the
model of admission and gender being conditionally independent, given department.
\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admission = FALSE))
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE,
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admission = FALSE))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}
Clearly, there is no gender bias in the departments except for department A: here,
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.
